---
title: "HWclass6"
author: "Jordan Prych (A17080226)"
date: "2025-01-24"
output: pdf_document
---

## HW in R Markdown

Output of original code: 

```{r}
#access bio3d database
library(bio3d)

#read in PDB structures 
 s1 <- read.pdb("4AKE")  # kinase with drug
 s2 <- read.pdb("1AKE")  # kinase no drug
 s3 <- read.pdb("1E4Y")  # kinase with drug
 
#takes the input of previous PDB structure and trims the file to a smaller subset of atoms, by sleecting chain A and elety
 s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
 s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
 s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
 
 #select for atom "b" from "atom" column (selecting beta factor from atom)
 s1.b <- s1.chainA$atom$b
 s2.b <- s2.chainA$atom$b
 s3.b <- s3.chainA$atom$b
 
#plotting beta factor
 plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor") 
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor") 
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

> Q6. How would you generalize the original code above to work with any set of input 
protein structures?

-Step 1 - read PDB code using `read.pdb()`. This function will read the file in PBD structures. The output accesses an-online PBD file and reads atom, seqres, helix, sheet, chain, and other variables pertaining to this protein.

-Step 2 - trim PDB structure to a smaller subset of atoms using `trim.pdb()`. This function outputs a trimmed PBD file regarded specified information from the arguments(chain and elety)

-Step 3 - select for atom "b" in chain A from "atom" column. This will select all "b" values in the "atom" column and output these values.

-Step 4 - plot values using `plotb3()` function. This will generate a scatter plot of the beta factor. 

```{r}
#access bio3d database 
library(bio3d)
#input to function is protein PDB file to read  
plot.pdb <- function(pdb.protein) {
#read in PDB structures 
   s <- read.pdb(pdb.protein)
#takes the input of previous PDB structure and trims the file to a smaller subset of atoms, by selecting chain A and elety
  s.chainA <- trim.pdb(s, chain="A", elety="CA")
#select for atom "b" from "atom" column (selecting beta factor from atom)
  s.b <- s.chainA$atom$b
#plotting beta factor
  plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor")
}
  
```

Testing generated function: 

```{r}
plot.pdb("1AKE")
```

```{r}
plot.pdb("4AKE")
```

```{r}
plot.pdb("1E4Y")
```